---
title: "P16"
author: "Nadeem Hassan"
---

```{r active="", eval=FALSE}
# BEGIN ASSIGNMENT 
```
The Fibonacci sequence is calculated as follows: 

F(x) = F(x - 1) + F(X - 2)

where F(1) = 1 and F(2) = 1, so you only have to start calculating from x = 3. For example, the first 10 numbers in the Fibonacci sequence are:

1 1 2 3 5 8 13 21 34 55

Calculate the `n`th Fibonacci numbers using a function with a for loop inside it. The autograder will check if there's a function called "`fibonacci`" (R is case sensitive) that takes an argument "`n`". It will also check that `fibonacci(3)` gives a 2, `fibonacci(4)` gives a 3, etc.

- Make sure your function is called `fibonacci` and takes one argument named `n`
- Make sure it returns *one* number (not the whole sequence). 
- Do not worry about making sure it works for n = 1 or n = 2, and don't worry about any other sort of error checking. 
  - Error checking is good practice for other functions, but won't affect your submission here.

```{r p16, error=TRUE}
# YOUR CODE HERE
fibonacci <- function(n) {
  if (n == 1 || n == 2) return(1)
  
  a <- 1
  b <- 1
  current <- 0
  
  for (i in 3:n) {
    current <- a + b
    a <- b
    b <- current
  }
  
  return(current)
}

fibonacci(n = 10)
```
```{r}
. = ottr::check("tests/fibby.R")
```

















It is well known that the ratio between [two consecutive Fibonacci numbers](https://en.wikipedia.org/wiki/Golden_ratio#Relationship_to_Fibonacci_and_Lucas_numbers) approaches the golden ratio:
$$
\phi = \frac{1 + \sqrt{5}}{2} = 1.618033988749
$$

Show this! Use a loop of some sort to calculate the ratio of $F(n) / F(n - 1)$ until we're "close enough", i.e. $|F(n) / F(n - 1) - \phi| < \epsilon$ for some small epsilon.

In other words, find `n` such that `abs(fibonacci(n) / fibonacci(n - 1) - golden_ratio) < 1e-100`. The autograder will search for an object named "`n`" with the correct answer.

(Note that `1e-100` is computer notation for 10^-1000, that is, a zero, then a decimal, then 99 zeros and then finally a 1.)

```{r gr, error=TRUE}
golden_ratio <- (1 + sqrt(5)) / 2
n <- 2

# Write some code to find the smallest n such that abs(fibonacci(n) / fibonacci(n - 1) - golden_ratio) < 1e-100
# YOUR CODE HERE
while (abs(fibonacci(n) / fibonacci(n - 1) - golden_ratio) >= 1e-100) {
  n <- n + 1
}
n
```
```{r}
. = ottr::check("tests/golden_ratio.R")
```

Fun fact: because of this fact, consecutive Fibonacci numbers can be used to convert from kilometers to miles! 

- 1 2
  - 1 mile is approximately 2 kilometers (1.609km, to be precise)
  - 2 kilometers is approximately 1 mile (1.247mi, to be precise)
- 2 3
  - 2mi is approx 3km (3.219)
  - 3km is approx 2mi (1.864)
- 3 5
  - 3mi is approx 5km (4.82)
  - 5km is approx 3mi (3.11)
- 5 8
- 8 13
- 13 21
  - 13mi is approx 21km (20.92)
  - 21km is approx 13mi (13.04)

Why is this true? Because 1mi / 1km = 1.608, which is close to the Golden Ratio (1.618). 

Next time you're travelling in the US, just open up R and write a while loop to find Fibonacci numbers that match the distance you need! (Or just trust your GPS, but I like the R approach.)

```{r active="", eval=FALSE}
# END ASSIGNMENT 
```
